Search techniques for gaussian mixture modelling with Minimum Message Length, with applications to the origin of our Galaxy

### 2018-10-02 article/ms.tex

\documentclass{elsarticle}
\journal{Artificial Intelligence Journal}
\usepackage{bm,url,amsmath,amssymb}
%\usepackage{boldsymbol}
\usepackage{color,colortbl} % for \todo command
\usepackage{verbatim} % for \begin{comment}'ing out sections.
\usepackage{algorithm,algorithmicx}
\usepackage{algpseudocode}

% For revision history
\IfFileExists{vc.tex}{\input{vc.tex}}{
    \newcommand{\githash}{UNKNOWN}
    \newcommand{\giturl}{UNKNOWN}}

@@

% Surveys
\newcommand{\apogee}{\acronym{APOGEE}}
\newcommand{\ges}{\acronym{GES}}
\newcommand{\hermes}{\acronym{HERMES}}
\newcommand{\galah}{\acronym{GALAH}}
\newcommand{\fourmost}{\acronym{4MOST}}
\newcommand{\weave}{\acronym{WEAVE}}
\newcommand{\gaia}{\project{Gaia}}
\newcommand{\Cspace}{$\mathcal{C}$-space }
\newcommand{\Cspaceno}{$\mathcal{C}$-space}
% Common terms
\newcommand{\vect}[1]{\boldsymbol{\mathbf{#1}}}
\renewcommand{\vec}[1]{\vect{#1}}

@@

@@

%% Group authors per affiliation:
\author[fit]{Aldeida Aleti\corref{mycorrespondingauthor}}
\cortext[mycorrespondingauthor]{Corresponding author}
\ead{aldeida.aleti@monash.edu}
\author[moca,fit]{Andy Casey}
\author[moca]{John Lattanzio}
\author[fit]{David L. Dowe}
\address[moca]{School of Physics and Astronomy, Monash University, Melbourne, Clayton VIC 3800, Australia}
\address[fit]{Faculty of Information Technology, Monash University, Melbourne, Clayton VIC 3800, Australia}

@@

        \begin{keyword}
            Search\sep Minimum Message Length\sep MML\sep mixture models\sep Galactic Archaeology
        \end{keyword}
    \end{frontmatter}
\section{Introduction} 
\label{sec:introduction}

The contributions of this paper are:
\begin{itemize}
\item Solving the galactic archaeology (GA) problem using mixture modelling
% bear in mind or don't forget that GA will mean `genetic algorithms' to many readers
\item Minimum Message Length (MML) formulation
\item 2 search methods
\end{itemize}
\section{Galactic Archaeology}
% "Galaxy produces ‘molecular forests’"
% https://CosmosMagazine.com/space/galaxy-produces-molecular-forests
Galactic Archaeology~\cite{freeman2002new} aims at reconstructing the history 
of our Galaxy by identifying ancient star formation events. Most stars are 
born in clusters (of order $10^0$ to $10^6$ stars at a time), and all approximately
share the same chemical composition of the gas cloud that they form from.
Stars in clusters often do not remain gravitationally bound, and disperse 
throughout the disk of our Galaxy within about $10^9\,{\rm yr}$. %citation needed.
As a consequence, where a star is now does not reflect its formation site, 
complicating our understanding of galaxy formation and evolution. However, the
observable chemical composition of a star remains largely the same throughout
that star's lifetime, providing an immutable marker (a chemical tag, or
`fingerprint') that carries with it information about where and when star
formed.

@@

Here $\prior{\vec{\theta}}$ is the prior, $Q$ is the number of free parameters
in the model, $\kappa(Q)$ is a function to approximate the lattice quantisation
constant for $Q$ free parameters\cite{ConwaySloane1984}, and $|\fisher{\vec{\theta}}|$
is the determinant of the \emph{expected} Fisher information matrix (the second
order partial derivatives of the negative log-likelihood function $-\log\likelihood\left(\vec{\data}|\vec{\theta}\right)$).
A common that arises after listing the components above is that although we
have expliticly encoded the model parameters, we have not encoded `the model'
in the sense that we have not described how these encoded parameters relate 
to each other. In a sense, we do not describe to the receiver that these
encoded parameters refer to a mixture of Gaussians, or what we mean when we
refer to the term `Gaussian'. This information is implicit, and justified
if we are evaluating and comparing models within the same class (e.g., gaussian 
mixture models). For example, it is analogous to writing the probability of
some data given a model as $p(\vec{\data}|\vec{\theta})$ rather than what is 
actually implied: $p(\vec{\data}|\vec{\theta},\vec{\mathcal{M}})$, the 
probability of the data given the model parameters $\vec{\theta}$ and class of 
model $\vec{\mathcal{M}}$. Whatever the message length required to encode to 
class of model $\vec{\mathcal{M}}$, under consideration, it would be a constant 
added to the message length regardless of the number of components in our mixture.

In the following subsections we provide expressions for the message length of
each component listed above, and our prior beliefs, before describing our
full objective function.

@@

We use an improper uniform prior on the mean $\vecmean$ for all mixtures such
that 
    $\mathcal{U}(\vecmean) = (-\infty, +\infty)$.
However,  the MML principle requires proper priors. In practice we make 
this prior proper by assuming the bounds on $\prior{\vecmean}$ are very large, but 
we do so without actually specifying the value of the bounds because they add 
only a constant value to the total message length\todo{\cite{someone}}.
We adopt a conjugate inverted Wishart prior for the covariance matrix of an
individual mixture \cite[Section 5.2.3]{Schafer_1997}:

@@

%\begin{eqnarray}
%I_K & = & K\log{2} % I_k
%    + \frac{(K - 1)}{2}\log{N} - \frac{1}{2}\sum_{k=1}^{K}\log{w_k} - \log{|\Gamma(K)|} % I_w
%    +\mathcal{L}(\data|\vectheta) \nonumber \\
%    & -& DN\log{y_{err}} % Likelihood
%   +  \frac{1}{2}\sum_{k=1}^{K}\left[\frac{D(D+3)}{2}\log{{Nw_k}} - (D + 2)\log{|\veccov_k|} - D\log{2}\right]\nonumber \\% I_t
%    &-& \frac{Q}{2}\log(2\pi) + \frac{\log{Q\pi}}{2} % lattice, minus the lattice of part 2
%%\end{eqnarray}

%\noindent{}where $Q = \frac{1}{2}DK(D + 3) + K - 1$, the number of free parameters, and
%$\mathcal{L}$ is the log-likelihood of a multivariate Gaussian distribution.

While we can evaluate the objective function $I(\vec{\data}|\vec{\psi})$, 
where for the sake of brevity we denote $\vec{\psi} = \{K, N, D, \vec{\weight}, 
\vecmean, \veccov\}$, and we can meaningfully compare models with different 
numbers of components or parameter values, we require a search strategy in 
order to efficiently find the global solution: the minimum message length 
over a class of finite mixture models (where $N > K$). Before we introduce and 
compare search strategies, let us first make some comments and quantitative 
descriptions on how the message length $I$ will vary with the model parameters. 
Specifically, we will re-arrange Equation \ref{eq:objective_function} to be a 
sum of three smaller functions, and we will derive limits and predictions for 
those quantites. In an ideal scenario, those limits and predictions would be 
able to provide us with an estimate of the message length for a Gaussian 
mixture before computing the mixture (i.e., before performing 
expectation-maximization.We begin by re-casting Eq. \ref{eq:objective_function} 
as
\begin{eqnarray}
I(\vec{\data}|\vec{\psi}) = I_{1}(K, N, D, \vec{\weight}) 
                          + I_{2}(K, N, \veccov)
                          + I_{3}(\vec{\psi}) \quad ,
\end{eqnarray}
\noindent{}where:
\begin{eqnarray}
    I_{1} &=& \left(1-\frac{D}{2}\right)K\log{2} + \log{\Gamma(K)} + \frac{1}{4}\left[2\left(K - 1\right) + KD\left(D+3\right)\right]\log{N} \nonumber \\
          &+& \frac{1}{2}\left(\log{Q\pi} - Q\log{2\pi}\right) + \left(\frac{D(D+3) - 2}{4}\right)\sum_{k=1}^{K}\log{\weight_k} \todo{+ \kappa} \quad , \label{eq:I1} \\
    I_{2} &=& -\left(1 + \frac{D}{2}\right)\sum_{k=1}^{K}\log{|\veccov_k|} \quad \textrm{and},\\
    I_{3} &=& -\log\likelihood\left(\vec{\data}|\vec{\theta}\right) - DN\log{\mathcal{E}} \quad .
\end{eqnarray}
The nomenclature for $I_1$, $I_2$ and $I_3$ bear no relation to the two-part 
message that we introduced earlier to describe message length. Rather we are
simply re-arranging the objective function because, as we will show, there are
useful theoretical bounds and predictions that we can make for each part of
this re-arranged expression.

@@

\begin{enumerate}
    \item The weights are distributed uniformly such that $w_k = 1/K$ and
          \begin{equation}
                \sum_{k=1}^{K}\log\weight_k = \sum_{k=1}^{K}\log{\frac{1}{K}} =-K\log{K} \quad .
          \end{equation}
    \item The weights are distributed such that $K - 1$ components each 
          describe just two data points $w_{1..K-1} = 2/N$, and the remaining 
          component describes all other data points $w_{K} = 1 - 2(K-1)/N$
          as $\sum_{k=1}^{K}\weight_k = 1$. At this extreme,
          \begin{equation}
                \sum_{k=1}^{K}\log\weight_k = 2(K-1)\log\frac{2}{N} 
                		    + \log\left(1 - \frac{2(K - 1)}{N}\right)
          \end{equation}
          \noindent{}and by making use of the identity 
            $\log{(a-b)} = \log{a} + \log{(1 - \frac{b}{a})}$:
          \begin{equation}
                \sum_{k=1}^{K}\log\weight_k = 2K\left(\log{2} - 1\right) - 2\log{2} + \log{N} + \log{\left(N - 2K + 2\right)} \quad .
          \end{equation}
\end{enumerate}

\noindent{}Thus, for any finite mixture of $K$ Gaussian components we have 
derived the bounds
\begin{eqnarray}
\sum_{k=1}^{K}\log\weight_k & \geq & -K\log{K} \nonumber \\
\sum_{k=1}^{K}\log\weight_k & \leq & 2K\left(\log{2} - 1\right) - 2\log{2} + \log{N} + \log{\left(N - 2K + 2\right)} \nonumber
\end{eqnarray}
\noindent{}which gives us analytic expressions for the upper and lower value 
of $I_1$, regardless of the values of $\vec{\weight}$, for any mixture, and
bounds on $I_{1}$ can be derived from just $\{K, N, D\}$.

@@

\section{Search Methods}

@@

\begin{eqnarray}
\Delta{}I_{K+1} &-& I_K \approx \log{2} % \Delta{}I_k
    + \frac{1}{2}\log{N} - \log{K} - \frac{1}{2}\left(\sum_{k=1}^{K+1}\log{w_k^{(new)}} - \sum_{k=1}^{K}\log{w_k^{(old)}}\right) \nonumber \\ % \Delta{}I_w
& +& \mathcal{L}^{(new)} - \mathcal{L}^{(old)} \nonumber \\ % likelihood
& +& \frac{1}{2}\left[\frac{D(D+3)}{2}\left(\sum_{k=1}^{K+1}\log{Nw_k^{(new)} - \sum_{k=1}^{K+1}\log{Nw_k^{(old)}}} \right) \right.\nonumber \\
&-& \left.\left(D+2\right)\left(\sum_{k=1}^{K+1}\log{|\veccov_k|^{(new)}} - \sum_{k=1}^{K+1}\log{|\veccov_k|^{(old)}}\right)\right] \nonumber \\
& +& \frac{\log(2\pi)}{2}(-\frac{1}{2}D(D+3) + 2K  - 1) + \frac{\pi}{2}\left(\log{(K + 1)} - \log{K}\right)
\end{eqnarray}

\begin{algorithm}[H]
\caption{The MML$_{K+D}$ Method.}
\label{alg:MMLKD}
\begin{algorithmic}[1]
\State {\bf \sc input:} $K$ \Comment{Number of initial components.}
\State current $\leftarrow$ $K$ component mixture
\While{TRUE}
\EndWhile
\State {\bf \sc output:} 
\end{algorithmic}
\end{algorithm}

>>>>

\documentclass{elsarticle}
\journal{Artificial Intelligence Journal}
\usepackage{bm,url,amsmath,amssymb}
%\usepackage{boldsymbol}
\usepackage{minted}
\usepackage{color,colortbl} % for \todo command
\usepackage{verbatim} % for \begin{comment}'ing out sections.
\usepackage{algorithm,algorithmicx}
\usepackage{algpseudocode}

% For revision history
\newcommand{\vcpath}{vc.tex}

\IfFileExists{\vcpath}{\input{\vcpath}}{
	\newcommand{\giturl}{UNKNOWN}
	\newcommand{\gitslug}{UNKNOWN}
	\newcommand{\githash}{UNKNOWN}
	\newcommand{\gitdate}{UNKNOWN}
	\newcommand{\gitauthor}{UNKNOWN}
}

@@

% Surveys
\newcommand{\apogee}{\acronym{APOGEE}}
\newcommand{\ges}{\acronym{GES}}
\newcommand{\hermes}{\acronym{HERMES}}
\newcommand{\galah}{\acronym{GALAH}}
\newcommand{\fourmost}{\acronym{4MOST}}
\newcommand{\weave}{\acronym{WEAVE}}
\newcommand{\gaia}{\project{Gaia}}
\newcommand{\Cspace}{$\mathcal{C}$-space }
\newcommand{\Cspaceno}{$\mathcal{C}$-space}

% Common terms
\newcommand{\vect}[1]{\boldsymbol{\mathbf{#1}}}
\renewcommand{\vec}[1]{\vect{#1}}

@@

\newcommand{\bashcommand}[1]{\noindent\texttt{#1}\\}

@@

%% Group authors per affiliation:
\author[fit]{Aldeida Aleti\corref{mycorrespondingauthor}}
\cortext[mycorrespondingauthor]{Corresponding author}
\ead{aldeida.aleti@monash.edu}
\author[moca,fit]{Andrew R. Casey}
\author[moca]{John C. Lattanzio}
\author[fit]{David L. Dowe}
\address[moca]{School of Physics and Astronomy, Monash University, Melbourne, Clayton VIC 3800, Australia}
\address[fit]{Faculty of Information Technology, Monash University, Melbourne, Clayton VIC 3800, Australia}

@@

        \begin{keyword}
            Search\sep Minimum Message Length\sep MML\sep gaussian mixture models\sep Galactic Archaeology
        \end{keyword}
    \end{frontmatter}
\section{Introduction} 
\label{sec:introduction}

The contributions of this paper are:
\begin{itemize}
\item Solving the galactic archaeology (GA) problem using mixture modelling
% bear in mind or don't forget that GA will mean `genetic algorithms' to many readers
\item Minimum Message Length (MML) formulation
\item 2 search methods
\end{itemize}
\section{Galactic Archaeology}
% "Galaxy produces ‘molecular forests’"
% https://CosmosMagazine.com/space/galaxy-produces-molecular-forests
Galactic Archaeology~\cite{freeman2002new} aims at reconstructing the history 
of our galaxy by identifying ancient star formation events. Most stars are 
born in clusters (of order $10^0$ to $10^6$ stars at a time), and all approximately
share the same chemical composition of the gas cloud that they form from.
Stars in clusters often do not remain gravitationally bound, and disperse 
throughout the disk of our galaxy within about $10^9\,{\rm yr}$. %citation needed.
As a consequence, where a star is now does not reflect its formation site, 
complicating our understanding of galaxy formation and evolution. However, the
observable chemical composition of a star remains largely the same throughout
that star's lifetime, providing an immutable marker (a chemical tag, or
`fingerprint') that carries with it information about where and when star
formed.

@@

Here $\prior{\vec{\theta}}$ is the prior, $Q$ is the number of free parameters
in the model,
\begin{eqnarray}
	Q = \frac{KD(D+3)}{2} + K - 1 \quad ,
\end{eqnarray}
\noindent{}$\kappa(Q)$ is a function to approximate the lattice quantisation
constant for $Q$ free parameters\cite{ConwaySloane1984}, and $|\fisher{\vec{\theta}}|$
is the determinant of the \emph{expected} Fisher information matrix (the second
order partial derivatives of the negative log-likelihood function $-\log\likelihood\left(\vec{\data}|\vec{\theta}\right)$).
A common that arises after listing the components above is that although we
have expliticly encoded the model parameters, we have not encoded `the model'
in the sense that we have not described how these encoded parameters relate 
to each other. In a sense, we do not describe to the receiver that these
encoded parameters refer to a mixture of Gaussians, or what we mean when we
refer to the term `Gaussian'. This information is implicit, and justified
if we are evaluating and comparing models within the same class (e.g., gaussian 
mixture models). For example, it is analogous to writing the probability of
some data given a model as $p(\vec{\data}|\vec{\theta})$ rather than what is 
actually implied: $p(\vec{\data}|\vec{\theta},\vec{\mathcal{M}})$, the 
probability of the data given the model parameters $\vec{\theta}$ and class of 
model $\vec{\mathcal{M}}$. Whatever the message length required to encode to 
class of model $\vec{\mathcal{M}}$, under consideration, it would be a constant 
added to the message length regardless of the number of components in our mixture.

In Appendix \ref{app:mml_derivation} we define our priors and describe the 
individual contributions to the mixture message length. For the sake of brevity we
denote the full set of model parameters as $\vec{\psi} = \{K, \vec{\weight}, \vecmean, \veccov\}$.
The total message length $I(\vec{\data}|\vec{\psi}, N, D)$ to model some $N \times D$ dimensional data $\vec{\data}$ with a mixture of multivariate Gaussian components is given by:
\begin{eqnarray}
	I(\vec{\data}|\vec{\psi}, N, D) &=& -\log\mathcal{L}\left(\vec{\data}|\vectheta\right) - \frac{(D + 2)}{2}\sum_{k=1}^{K}\log{|{\cov}_k|} \nonumber \\
	&& \cdots\ + \left(\frac{D(D+3)}{4} - \frac{1}{2}\right)\sum_{k=1}^{K}\log{\weight_k} + K\left(1-\frac{D}{2}\right)\log{2} \nonumber \\ && \cdots\ + \log\Gamma{(K)} + \frac{1}{2}\left(\log{Q\pi} - Q\log{2\pi}\right) + \frac{Q}{2}\log{N} \,\,\,\, . 
	\label{eq:objective_function}
\end{eqnarray}

\subsection{Expectation-Maximization}

\todo{Derivations and updates to the mixture parameters through E-M}

\subsection{Theoretical lower bounds on the message length of future mixtures}

Knowing how some of the model parameters are distributed allows us to derive theoretical
bounds on the message length for individual components. These bounds allow us to describe
the lower and upper bounds on \emph{some components of} the message length of a Gaussian mixture
model, before estimating the model parameters (or running expectation-maximisation). For the
purposes of this work we will describe such mixtures as \emph{future mixtures} as they have
not been trialled yet.

In Appendix \ref{app:mml_limits} we describe four useful bounds. The first two are lower and upper bounds on the   message length to encode the sum of the log of the weights $\sum_{k=1}^{K}\weight_k$
for any future mixture. The third is an approximation to the lower bound of the negative log-likelihood $-\log\mathcal{L}\left(\vec{\data}|\vec{\theta}\right)$, and the fourth is an approximation on the message length required to encode
the sum of the log of the determinant of the covariance matrices $\sum_{k=1}^{K}\log|\cov_k|$. These are the  most important components in the objective function described
in Equation \ref{eq:objective_function}, because their values are not known until some set of model parameters $\vec{\psi}$ is trialled -- and the optimal set of model parameters is not known \emph{a priori} -- and all other terms in Equation \ref{eq:objective_function} are deterministic, depending only on $K$, $N$, and $D$.
While we do provide both a lower and upper bound on the message length to encode the weights, no such upper bound exists on the
negative log-likelihood because this could be arbitrarily high for an increasingly poor estimate of the mixture
parameters. Some upper limit likely exists for $\sum\log|\cov_k|$, but that limit would depend on how the data are separated. Nevertheless, it is the lower bound on these quantities that is most useful for optimisation.

Empirically we found that when the dimensionality of the data $D$ is small, these theoretical limits can efficiently describe a lower bound to the message length for any given Gaussian mixture of $K$ components. This allows, in some situations, to efficiently place a useful upper limit on the maximum number of components $K$, $K_{max}$ (useful in that $K_{max} << N$).

\subsection{Predicting the message length of future mixtures}

The objective function in Equation \ref{eq:objective_function} remains valid across a class of 
Gaussian mixture models where $N > K$. However, we still require a strategy to minimise this
objective function over all possible mixture parameters $\vec{\psi}$. Here we describe methods to predict
the \emph{minimum} message length of future (untrialled) mixtures, which we will then use in conjunction 
with search strategies (Section \ref{sec:search}).

% TODO: Predicting the sum of the log of the weights.

% TODO: Predicting the negative log-likelihood.

% TODO: Predicting the sum of the log of the determinant of the covariance matrices.

% Combining it together.

\section{Search Strategies} \label{sec:search}

We introduce and evaluate two search strategies that make use of the objective function in Equation \ref{eq:objective_function}. These search strategies are compared against an existing state-of-the-art method, which we describe in Section \ref{sec:experimental-design}.

\subsection{The Message-Breaking Method} \label{sec:mbm}
The first strategy, \emph{The Message-Breaking Method}, does not make any predictions on
the message length of future mixtures. Instead it starts from 
a $K=1$ mixture and iteratively splits the component with the largest message length until 
the total message length no longer improves (Algorithm~\ref{alg:MBM}). We evaluate the message length of each component
by re-ordering Equation \ref{eq:objective_function} and weighting the negative log-likelihood
$-\log{\mathcal{L}}(\vec\data|\vectheta)$ by the $N \times K$ responsibility matrix $\mathcal{R}$ with entries
\begin{eqnarray}
	\mathcal{R}_{nk} = \frac{\weight_{k}f(\vec\data_n|\vecmean_k,\veccov_k)}{\sum_{k=1}^{K}\weight_{k}f(\vec\data_n|\vecmean_k,\veccov_k)}
\end{eqnarray}
\noindent{}such that the message length of the $k$th component is
\begin{eqnarray}
	I_k &=& -\sum_{n=1}^{N}\mathcal{R}_{nk}\log{\weight_{k}f(\data_{n}|\vecmean_k,\veccov_k)} -\frac{(D + 2)}{2}\log{|\cov_k|} \nonumber \\
		&& \cdots\ + \left(\frac{D(D+3)}{4}-\frac{1}{2}\right)\log{\weight_k} + \left(1 - \frac{D}{2}\right)\log{2} \nonumber \\
		&& \cdots + \frac{1}{K}\left[\log\Gamma(K) + \frac{1}{2}\left(\log{Q\pi} - Q\log{2\pi}\right) + \frac{Q}{2}\log{N}\right]
\end{eqnarray}
\noindent{}and the total message length remains the sum of the message lengths of the components
\begin{eqnarray}
	I(\vec\data|\vec\psi, N, D) = \sum_{k=1}^{K}I_{k} \quad .
\end{eqnarray}

The component-wise calculation of the message length is performed as part of the 
expectation step of the expectation-maximisation algorithm, adding negligible computational overhead.

When splitting the component with the largest message length, we split this component into two child components. Initially we 
set the means of the child mixtures as positive and negative eigenvalues along the principal
eigenvector of the component covariance matrix. A matrix of child responsibilities is then
calculated based on the Euclidean distance between each datum and the child means (i.e., hard
association), which we use to estimate the covariance matrices and weights of the child
components. After the parameters of the child components are initialised, we run \emph{partial}
expectation-maximization to iteratively update the parameter estimates of the child components. 
By \emph{partial} expectation-maximization we refer to the process of keeping the parameters of
all other components fixed, and only updating the estimates of the parameters of the child
components \citep{kasarapu2015minimum}. After partial E-M has converged, we integrate the child components with the rest of the mixture and run full E-M before the \textsc{Split} operation (line \ref{line:mbm-split} of Algorithm~\ref{alg:MBM}) is considered complete.

One advantage to the Message-Breaking Method is that the total number of E-M steps is relatively few because all partial E-M steps are performed from a near-optimal initialisation point. One possible disadvantage to the Message-Breaking Method is that there may be mixtures where the \textsc{Split} operation increases the message length by $\epsilon$, but a subsequent \textsc{Split} operation would \emph{decrease} the total message length by $>\epsilon$. As a result, the algorithm would converge to a local minima, not the global one. However, as we show in Section \ref{sec:results}, this hypothetical situation was not observed in any of the $\todo{5000}$ generated datasets.

\begin{algorithm}[H]
\caption{The Message-Breaking Method.}
\label{alg:MBM}
{\bf Input:} $\vec{\data}$ \Comment{The $N \times D$ dimensional data.}\\
{\bf Output:} $\vec{\psi}$ \Comment{The optimal mixture parameters.} 
\begin{algorithmic}[1]
\State current $\leftarrow$ one component mixture
\While{TRUE}
\State $m$ $\leftarrow$ \Call{ComponentMessageLength}{current}
\State $i$ $\leftarrow$ \Call{ArgMax}{$m$} 
\State perturbation $\leftarrow$ \Call{Split}{current, components[i]} \label{line:mbm-split}
\State $\Delta{}I \leftarrow$ \Call{MessageLength}{perturbation} - \Call{MessageLength}{current} 
\If{$\Delta I < 0$} \Comment{Check for improvement.}
\State current $\leftarrow$ perturbation
\State {\bf \sc Continue}
\EndIf
\State {\bf \sc Break} \label{line:end}
\EndWhile
\end{algorithmic}
\end{algorithm}

\subsection{Component-Jumping Method}

The \emph{Component-Jumping Method} seeks to find the mixture parameters $\vec\psi$ that produce the minimum message length over the class of finite mixtures $\vec{\mathcal{M}}$ in the fewest number of model evaluations.

\begin{algorithm}[H]

\algblock[TryCatchFinally]{try}{endtry}
\algcblock[TryCatchFinally]{TryCatchFinally}{finally}{endtry}
\algcblockdefx[TryCatchFinally]{TryCatchFinally}{catch}{endtry}
	[1]{\textbf{catch} #1}
	{\textbf{end try}}

\caption{The Component-Jumping Method.}
\label{alg:component-jumping-method}
{\bf Input:} $\vec\data$, $K_{i}$, $I_\epsilon$ \\
{\bf Output:} $\vec\psi$ \Comment{The optimal mixture parameters.}
\begin{algorithmic}[1]
\State s $\leftarrow (K_{i})^{-1}\log_{10}{N}$ \Comment{Initial step size in $K$.}
\State v $\leftarrow [0, 1, ..., K_{i}]$ 
\State $K_{inits} \leftarrow$ \Call{Power}{10, vs} \Comment{Set initial $K$ values to trial.}
\State list $K_{t}$ \Comment{Create a list to save trialled $K$ values.}
\For{$k\in K_{inits}$} 
  \State $K_{t}$.append($k$) \Comment{Remember that we trialled this $K$ value.}
  \try 
	  \State mix $\leftarrow$ \Call{KMeans$^{++}$}{k} \Comment{Initialise with K-means++.}
	  \State mix $\leftarrow$ \Call{ExpectationMaximization}{mix}
  \catch
  	  \State {\bf \sc Break} \Comment{Do not trial any higher $K$ values.}
  \finally 
      \State mixture[k] $\leftarrow$ mix 
      \State state[k] $\leftarrow$ \Call{MessageLength}{mixture[k]}
  \endtry
\EndFor
\State converged $\leftarrow$ FALSE
\While{not converged}
	\State $K_{predict} \leftarrow [1, \cdots, 2\Call{Max}{K_{t}}]$ \Comment{Predict future message lengths.}
	\State $I_{predict} \leftarrow$ \Call{PredictMessageLength}{$K_{predict}$}
	% argsort.
	\State $K_{next} \leftarrow$ \Call{ArgSort}{$I_{predict}$} \Comment{Sort $K$ by predicted message length.}
	\For{$k\in K_{next}$}
		% Check if the mixture at K was good?
		\If{$k\in K_{t}$}
			\State converged $\leftarrow$ TRUE
			\State {\bf \sc Break}
		\Else
			\State $K_{t}$.append($k$) \Comment{Remember that we trialled this $K$ value.}
	   	  \try 
		    \State mix $\leftarrow$ \Call{KMeans$^{++}$}{k} \Comment{Initialise with K-means++.}
	   	    \State mix $\leftarrow$ \Call{ExpectationMaximization}{mix}
		  \catch
  	        \State {\bf \sc Continue} \Comment{Move on to next $K$ trial.}
		  \endtry 
		\EndIf
	\EndFor
\EndWhile
\end{algorithmic}
\end{algorithm}

The different search schemes are validated using the Kolmogorov-Smirnov (KS) non-parametric test~\citep{Pettitt77S}, which checks for a statistical significance in the difference between the performances of the algorithms. The 50 runs are submitted to the KS analysis with a null hypothesis of no difference between the performances of the different search methods.  \andy{Huh?}

\section{Appendices}
\subsection{The MML objective function}
\label{app:mml_derivation}
In the following subsections we provide expressions for the message length of
each component listed above, and our prior beliefs, before describing our
full objective function.

@@

We use an improper uniform prior on the mean $\vecmean$ for all mixtures such
that 
    $\mathcal{U}(\vecmean) = (-\infty, +\infty)$.
However,  the MML principle requires proper priors. In practice we make 
this prior proper by assuming the bounds on $\prior{\vecmean}$ are very large, but 
we do so without actually specifying the value of the bounds because they add 
only a constant value to the total message length.
We adopt a conjugate inverted Wishart prior for the covariance matrix of an
individual mixture \cite[Section 5.2.3 of][]{Schafer_1997}:

@@

\section{Bounds on components of the message length}

@@

\begin{enumerate}
    \item The weights are distributed uniformly such that $w_k = 1/K$ and
          \begin{equation}
                \sum_{k=1}^{K}\log\weight_k = \sum_{k=1}^{K}\log{\frac{1}{K}} =-K\log{K} \quad .
          \end{equation}
    \item The weights are distributed such that $K - 1$ components each 
          describe just two data points $w_{1..K-1} = 2/N$, and the remaining 
          component describes all other data points $w_{K} = 1 - 2(K-1)/N$
          as $\sum_{k=1}^{K}\weight_k = 1$. At this extreme,
          %\begin{equation}
          %      \sum_{k=1}^{K}\log\weight_k = (K-1)\log\frac{1}{N} + \log\left(1 - \frac{K - 1}{N}\right)
          %\end{equation}
          \begin{equation}
                \sum_{k=1}^{K}\log\weight_k = 2(K-1)\log\frac{2}{N} 
                		    + \log\left(1 - \frac{2(K - 1)}{N}\right)
          \end{equation}

          \noindent{}and by making use of the identity 
            $\log{(a-b)} = \log{a} + \log{(1 - \frac{b}{a})}$:
          \begin{equation}
                \sum_{k=1}^{K}\log\weight_k = 2K\left(\log{2} - 1\right) - 2\log{2} + \log{N} + \log{\left(N - 2K + 2\right)} \quad .
          \end{equation}
          %\begin{equation}
          %      \sum_{k=1}^{K}\log\weight_k = \log{(N-K+1)} - K\log{N} \quad .
          %\end{equation}
\end{enumerate}

\noindent{}Thus, for any finite mixture of $K$ Gaussian components we have 
derived the bounds
\begin{eqnarray}
\sum_{k=1}^{K}\log\weight_k & \geq & -K\log{K} \nonumber \\
\sum_{k=1}^{K}\log\weight_k & \leq & 2K\left(\log{2} - 1\right) - 2\log{2} + \log{N} + \log{\left(N - 2K + 2\right)} \nonumber
\end{eqnarray}
\begin{equation}
-K\log{K} \quad \leq \quad \sum_{k=1}^{K}\log\weight_k \quad \leq \quad (K - 1)\log{2} - K\log{N} + \log{(N - 2K + 2)}
\end{equation}

\noindent{}which gives us analytic expressions for the upper and lower value 
of $I_1$, regardless of the values of $\vec{\weight}$, for any mixture, and
bounds on $I_{1}$ can be derived from just $\{K, N, D\}$.

@@

\subsection{Reproducibility} \label{sec:reproducibility}

\noindent{}This project was developed in a public \texttt{git} repository that is 
accessible at:
\begin{center}
\giturl
\end{center}
\noindent{}This repository includes a well-tested \texttt{Python 3.6} implementation
of the algorithms described here, scripts to reproduce our results, and \LaTeX\ to compile this manuscript. This research can be reproduced in full using
the following commands in a modern terminal:\\\\
\bashcommand{git clone https://github.com/andycasey/gmmmml.git gmmmml}
\bashcommand{cd gmmmml/}
\bashcommand{git checkout \githash}
\bashcommand{./reproduce}

Reproducing these results will require at least \todo{X}\,Gb of free disk space
and \todo{Y}\,hours of compute time, mostly because of the extensive algorithm 
comparisons performed.

%---------
% COMMENTS
%---------

\begin{comment}
We employ two criteria for comparing the proposed search methods to the benchmark: the message length and the Kullback Leibler divergence~\cite{kullback1951information}.

\paragraph{Kullback Leibler divergence}
\end{comment}

\begin{comment}

@@

\begin{eqnarray}
\Delta{}I_{K+1} &-& I_K \approx \log{2} % \Delta{}I_k
    + \frac{1}{2}\log{N} - \log{K} - \frac{1}{2}\left(\sum_{k=1}^{K+1}\log{w_k^{(new)}} - \sum_{k=1}^{K}\log{w_k^{(old)}}\right) \nonumber \\ % \Delta{}I_w
& +& \mathcal{L}^{(new)} - \mathcal{L}^{(old)} \nonumber \\ % likelihood
& +& \frac{1}{2}\left[\frac{D(D+3)}{2}\left(\sum_{k=1}^{K+1}\log{Nw_k^{(new)} - \sum_{k=1}^{K+1}\log{Nw_k^{(old)}}} \right) \right.\nonumber \\
&-& \left.\left(D+2\right)\left(\sum_{k=1}^{K+1}\log{|\veccov_k|^{(new)}} - \sum_{k=1}^{K+1}\log{|\veccov_k|^{(old)}}\right)\right] \nonumber \\
& +& \frac{\log(2\pi)}{2}(-\frac{1}{2}D(D+3) + 2K  - 1) + \frac{\pi}{2}\left(\log{(K + 1)} - \log{K}\right)
\end{eqnarray}
\end{comment}

\begin{comment}
\subsection{Method of Moments}
% \alda{This is a copy paste from David's email with minor changes}
%
% Re the case of two (or more) classes with the same mean:
% In 2D, a simple case in point is two classes with the same mean
% such that $\sigma_{1, x} = \sigma_{2, y} >> \sigma_{2, x} = \sigma_{1, y}$.
% Andy said that this and higher dimensional stuff can be picked up
% by dropping to 1 dimension along the eigenvector of largest eigenvalue.
%
%
% I wonder also about the merits of dropping to 2 dimensions (if we have at least 2 attributes), being the eigenvectors of the two largest eigenvalues. 
%
% One could possibly check whether uniform radially directly from the original 2D projection without having to project it to a circle (i.e., it might be pretty easy to work out what the angle of each point would be post-transformation without having to do the transformation).  The 2D projection is 1D in the sense that we only care about the angle around the circle.  And then some test (e.g., Kolmogorov-Smirnov test) for whether data is uniform in radial distribution.  If not uniform, then perhaps grounds for splitting.
%
% Re looking at things in 1 dimension (not 2D followed by 1D [0, 2 pi) angle
% as above but rather projecting on eigenvector of largest eigenvalue), we
% discussed moments - e.g., \url{https://en.wikipedia.org/wiki/Normal_distribution}.
% So, 2nd, 4th, 6th and 8th moments of a true Gaussian are 1, 3, 15, 105.
%
